---
title: "Estimation of Long Run Effects in R"
author: "André Walter"
date: 2017-09-13
tags: ["ADL", "Dynamic Panels","R","Econometrics"]
categories:
  - Statistics
  - R
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

A lot of quantitative data in political science comes in the form of "Time-Series Cross-Sectional" (TSCS) datasets. Such datasets usually consist of a low number of cross-sectional units (e.g. states, countries) and higher number of time periods. Since the introduction of panel-corrected standard errors in political science in 1995 by Beck and Katz, political science scholars have discussed the merits and pitfalls of TSCS data. One controversial aspect is the inclusion of lagged dependent variables (LDVs) in estimation equations. Some have argued that LDVs supress the explanatory power of other predictors (see [Achen 2000](http://www.princeton.edu/csdp/events/Achen121201/) or [Plumper et al. 2005](http://onlinelibrary.wiley.com/doi/10.1111/j.1475-6765.2005.00230.x/abstract)). Others have stressed that the omission of LDVs leads to an considerable bias if the data generating process is dynamic ([Keele 2006](https://www.cambridge.org/core/journals/political-analysis/article/dynamic-models-for-dynamic-theories-the-ins-and-outs-of-lagged-dependent-variables/F4AB52C3E9964515825D1E6F20C9EA42)).

Without going into the details of this discussion, many scholars want to investigate how their results are affected by different model specifications such as the inclusion of LDVs (or reviewers require them to do so etc.). Since I'm using R for most of my quantitative work, I provide a guideline of how to estimate dynamic panels with TSCS data and, most importantly, obtain the long run effects of the predictors. 

To illustrate how it's done, I employ the Crime data from the plm package to estimate an autoregressive distributed lag model (Partial Adjustment). To do so, I use the DataCombine package to create a lagged dependent variable and the plm package for estimation. Crime is a data set with information on crime rates in North Carolina (1981-1987) and a few predictors. I use crime rates as dependent variable and the lag of crime rates as well as the number of police per capita (The lag is created by using the dplyr package). Our equation looks as follows:

$$y_{i,t} = \alpha_{i,t}+\gamma*y_{i,t-1}+\beta*polpc_{i,t}+\epsilon_{i,t}$$

```{r }
library(plm)
library(dplyr)
data(Crime)

Crime <- Crime %>% group_by(county) %>% mutate(lagcrm=lag(crmrte, na.rm =TRUE))

fit <- plm(crmrte ~ lagcrm+polpc, data = Crime, index = c("county","year"), model = "pooling")
summary(fit)
```


Lagcrm is significant with a point estimate of .88. The same goes for polpc with 0.65. Since all variables are in log, we can say that an one percent increase in police leads to a .65 percent increase in crime rates in the short run (i.e. more police p.c., more reported crimes). 

What is the long run effect? We can calculate the point estimate by simply by hand (again, partial adjustment model): 

$$y^*=\frac{\beta}{1-\gamma}$$

In our example, it's 

$$5.42=\frac{.65}{1-.88}$$

The point estimate shows that an one percent increase in police p.c. is associated with an 5.4 percent increase in crime rates - a larger change than the short run effect! However, the uncertainty of our point estimate is still missing.

There are two ways to deal with this problem. First, we can use instrumental variable (IV) estimation to obtain the standard errors ([Bewley 1979](http://www.sciencedirect.com/science/article/pii/0165176579900119)). Second, we can use the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html) ([Papke and Woodridge 2005](http://www.sciencedirect.com/science/article/pii/S0165176504003143)).

Which one should be used? It is a matter of convenience. In R, the plm package allows us to use IV estimation with cluster robust or panel corrected standard errors. Therefore, there wasn't any reason (for me) to work with the delta method. However, you might want to work with a postestimation fix for the standard errors that is not available for IV estimation (e.g. if you work with Stata instead of R and use commands such as xtpcse). Then the delta method comes in handy.

For IV (or 2SLS) estimation, we proceed as follows: In the first equation, we regress the differenced values of crime rates on the lag of crime rates:

$$\Delta y_{i,t}=\alpha_{i,t}+\gamma*y_{i,t-1}+\epsilon_{i,t}$$

In the second stage, we use then:


$$y_{i,t} = \alpha_{i,t}+\gamma*\Delta y_{i,t-1}+\beta*polpc_{i,t}+\epsilon_{i,t}$$

So, in R the long run effects can be estimated with IV estimation as follows: First, we use the dplyr package to generate our differenced values for crime rates. Then we estimate the long run effects with the plm function: 

```{r }
library(dplyr)

Crime <- Crime %>% group_by(county) %>% mutate(diff=crmrte-lag(crmrte, na.rm =TRUE))
fitlr <- plm(crmrte ~ diff+polpc | lagcrm+polpc, data = Crime, index = c("county","year"), model = "pooling")
summary(fitlr)

```

The point estimate for polpc is appro. 5.43. The result comes close to want we've calculated by hand. In addition, we've obtained the standard error for the estimate.

In order to get the same result using the delta method, we have to work with the car package.

```{r }
library(car)
dfitlr <- deltaMethod(fit,"polpc/(1-lagcrm)")
dfitlr

```
The delta method delivers the same result. Using the delta method requires less coding. However, presenting the results via regression tables or coefficient plots (with the full model) is less tedious with IV estimation.
